Containers and R---the Rockerverse and beyond

### 2019-11-30 manuscript.Rmd
some more edits

---
title: "Containers and R---the Rockerverse and beyond"
author:
  - name: Daniel Nüst
    affiliation: University of Münster
    address:
      - Institute for Geoinformatics
      - Heisenbergstr. 2
      - 48149 Münster, Germany
      - \href{https://orcid.org/0000-0002-0024-5046}{\includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} https://orcid.org/0000-0002-0024-5046}
    email: daniel.nuest@uni-muenster.de
  - name: Carl Boettiger
    #affiliation: -
    #email: -
  - name: Robrecht Cannoodt
    #affiliatin: -
    #email: -
  - name: Dirk Eddelbuettel
    affiliation: University of Illinois at Urbana-Champaign
    address:
      - Department of Statistics
      - Illini Hall, 725 S Wright St
      - Champaign, IL 61820, USA
      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0001-6419-907X}{https://orcid.org/0000-0001-6419-907X}
    email: dirkd@eddelbuettel.com
  - name: Mark Edmondson
    affiliation: IIH Nordic A/S, Google Developer Expert for GCP
    email: mark@markedmondson.me
  - name: Colin Fay
    affiliation: ThinkR
    address:
      - 5O rue Arthur Rimbaud
      - 93300 Aubervilliers, France
      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0001-7343-1846}{https://orcid.org/0000-0001-7343-1846}
    email: contact@colinfay.me
  - name: Ben Marwick
    #affiliation: -
    #email: -
  - name: Karthik Ram
    affiliation: Berkeley Institute for Data Science
    address:
      - University of California
      - Berkeley, CA 94720, USA
      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0233-1757}{https://orcid.org/0000-0002-0233-1757}
    email: karthik.ram@berkeley.edu
  - name: Noam Ross
    affiliation: EcoHealth Alliance
    address:
      - 460 W 34th St., Ste. 1701
      - New York, NY 10001, USA
      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0233-1757}{https://orcid.org/0000-0002-0233-1757}
    email: ross@ecohealthalliance.org
  - name: Nan Xiao
    affiliation: Seven Bridges Genomics
    address:
      - 529 Main St, Suite 6610
      - Charlestown, MA 02129, USA
      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0250-5673}{https://orcid.org/0000-0002-0250-5673}
    email: me@nanx.me
  - name: Lori Shepherd
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
      - Elm & Carlton Streets
      - Buffalo, New York, 14263, USA
      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-5910-4010}{https://orcid.org/0000-0002-5910-4010}
    email: lori.shepherd@roswellpark.org
  - name: Nitesh Turaga
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
    - Elm & Carlton Streets
    - Buffalo, New York, 14263, USA
    - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0224-9817}{https://orcid.org/0000-0002-0224-9817}
    email: nitesh.turaga@roswellpark.org
abstract: >
  The Rocker project provides widely-used Docker images for R across different application scenarios.
  This articles surveys describe downstream projects building upon Rocker.
  We also look beyond Rocker to other projects connecting containerisation with R.
  These use cases and the diversity of applications demonstrate the power of Rocker and containerisation for collaboration, effectivity, scalability, and transparency.
# blank footnote? https://tex.stackexchange.com/questions/170511/footnotes-without-numbering
output:
  rticles::rjournal_article:
    keep_tex: true
---

@@

The R community keeps growing: the number of new packages on CRAN keeps on rising, meetups, conferences, online courses and unconferences prosper, and the uptake in education and industry increases [REFs].
All this cements the role of R as the lingua franca of statistics, data visualisation, and computational research.
Coinciding with the rise of R was the advent of Docker [CITATION?] as a general tool for development, distribution, and deployment of software.
Combining both these topics, the _Rocker Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) was introduced in 2014 and provides a number of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) images for various use cases as described in \citet{RJ-2017-065}.
The considerable uptake and continued evolution of the Rocker suite of containers has lead to numerous projects extending or building upon Rocker images ranging from reproducible research to production deployments.
This article presents this _Rockerverse_ of projects.
It also introduces related activities connecting the R language and environment with other containerisation solutions.
The main contributions is a coherent picture of the current lay of the land of containers in, with, and for `R`.
This diversity includes demonstrators, early prototypes, and mature projects.

# Containerization and Rocker [NN]

<!--_Do we need a generic intro here?_ DE: Yes, and happy to take a stab.-->

Docker, an application and service provide by the eponymous company, has in just a few short years risen to prominence for development, testing, deployment and distribution of computer software.  While there are related approaches such as LXC or Singularity, Docker has become synomymous with "containerization"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operationg systems.   In doing so, Docker had recognised and validated the importance of one very important thread that had been emerging, namely virtualization. By allowing multiple concurrent applications or services to run concurrently on one host machine without any fear of interference between them, an important scalability feature had been realized.  But Docker improved on virtualization by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualization.  This makes for more efficient use of system resources and allowed another order of magnitude in terms of scalability of deployment. [Reference?]   While Docker makes use of Linux kernel features, its importance was large enough so that some required aspects of running Docker have been added to other operating systems to support Docker there more efficiently too. [Win10; reference?]

The key accomplishment of Docker is to make a "bundled" aggregation of software available to any system running Docker, without requiring much else from the host.  This is a rather attractive, and novel, proposition which has lead to widespread adoption and use of Docker in a variety of domains.  It provided also a natural match for "cloud deployment" which runs (or at least appears to run) "seamlessly" without much explicit reference to the underlying machine, architecture or operating system.  Using containers on cloud hosted servers is a natural fit also because these containers can be deployed with very littler in terms of dependencies on the host system.

For statistical computing and analysis centered around R, the Rocker Project has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}. The Rocker Project provides several lines of containeres spanning to from building blocks with R-release or R-devel to containers with RStudio Server and Shiny Server. Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN. [Maybe at a later point add forward reference to other topics building on Rocker described here.]

@@

**Microsoft R Open** ([MRO](https://github.com/nuest/mro-docker)) is an R distribution formerly known as Revolution R Open (RRO) before Revolution Analytics was acquired by Microsoft.
MRO is compatible with main R and its packages.
"It includes additional capabilities for improved performance, reproducibility, and platform support." \citep{microsoft_mro}
Most notably these capabilities are the MRAN repository a.k.a. CRAN Time Machine, which is enabled by default, and the (optional) integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread}.
<!-- MRAN is also used by versioned Rocker images. -->
MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images `nuest/mro` are available on Docker Hub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) and on GitHub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) respectively.
The images are inspired by the Rocker images and can be used much in the same fashion, effectively a drop-in replacement allowing users to quickly evaluate if the benefits of MRO + Intel® MKL apply to their use case.
Version-tagged images are provided for the latest bugfix release of recent R versions.
Extended license information about MKL is printed at every startup.

**Renjin** is an interpreter for the R language running on top of the [Java Virtual Machine](https://en.wikipedia.org/wiki/Java_virtual_machine) (JVM) providing full two-way access between Java and R code \citep{wikipedia_renjin_2018}.
It was developed to combine the benefits of R, such scripting and extension packages, with the JVM's advantages in the areas of security, cross-platform availability, and established position in enterprise settings.
R extension packages need to be specially compiled and are distributed via the Java package manager [Apache Maven](https://en.wikipedia.org/wiki/Apache_Maven), cf. [http://packages.renjin.org/packages](http://packages.renjin.org/packages) for available packages.
Packages are loaded on demand, i.e. at the first call to `library()`.
Not all R packages, especially one linking to binary libraries, are available, e.g. `rgdal` \footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.
There are no offical Docker images for Renjin, but community-maintained images for selected releases only are available under `nuest/renjin` on Docker Hub and GitHub at [https://hub.docker.com/r/nuest/renjin](https://hub.docker.com/r/nuest/renjin) and [https://github.com/nuest/renjin-docker](https://github.com/nuest/renjin-docker) respectively.
These images expose the command line interface of Renjin in a similar fashion as Rocker images and allow an easy evaluation of Renjin's suitability, but are not intended for production use.

**[pqR](http://www.pqr-project.org/)** is a "a pretty quick version of R".
_pqR_ fixes some opinionated issues in the R language and is the basis for experimental features, e.g. automatic diffentiation\footnote{ \href{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}}.
The source code development [on GitHub](https://github.com/radfordneal/pqR/) is a one man project and it does not provide any Docker images.
But especially disruptive approaches may contribute to the development of the R ecosystem, so the `nuest/pqr` images on Docker Hub and GitHub at [https://hub.docker.com/r/nuest/pqr/](https://hub.docker.com/r/nuest/pqr/) and [https://github.com/nuest/pqr-docker](https://github.com/nuest/pqr-docker) respectively.

**FastR** is "A high-performance implementation of the R programming language, built on GraalVM" (https://github.com/oracle/fastr).
It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R \footnote{ \href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
There are no offical Docker images provided, but `nuest/fastr` images and `Dockerfile` are independently maintained on [Docker Hub](https://cloud.docker.com/repository/docker/nuest/fastr/) and [GitHub](https://github.com/nuest/fastr-docker) respectively.

While the images presented in this section are far from being as vetted, stable, and widely used as any of the Rocker images, they demonstrate an important advantage of containerisation technology, namely the abiliy to transparently build portable stacks of open source software and make them easily accessible to users.
All different distributions are published under GPL licenses.
Since all of the different R distributions claim better performance as a core motivation, a comparision based on Docker images, potentially leveraging the [resource restriction mechanisms](https://docs.docker.com/config/containers/resource_constraints/) of Docker to level the playing field, seems useful future work.

@@

[_Bioconductor_](http://bioconductor.org/) is an open source, open development project for the analysis and comprehension of genomic data \citep{gentleman_bioconductor_2004}.
The project consists of 1741 R software packages as of August 15th 2019, as well as packages containing annotation or experiment data.
_Bioconductor_has a semiannual release cycle, each release is associated with a particular version of R.
Docker images allow availability of current and past versions of _Bioconductor_ for convenience and reproducibility.
_Bioconductor_ 'base' docker images are built on top of `rocker/r-ver` and `rocker/rstudio`.
_Bioconductor_ installs packages based on the R version, and therefore uses `rocker/rstudio` and `rocker/r-ver` version tagging.
_Bioconductor_ selects the desired version of R from Rocker, adds the BiocManager CRAN package for installing appropriate versions of _Bioconductor_ packages, and creates a _Bioconductor_ docker image with an informative tag (R\_version\_Bioc\_version).
The images are summarized on the _Bioconductor_ web site (https://bioconductor.org/help/docker/), maintained on GitHub ([https://github.com/Bioconductor/bioc_docker](https://github.com/Bioconductor/bioc_docker)), and available to the community through [DockerHub](https://hub.docker.com/).
Past and current combinations of R and _Bioconductor_ are therefore accessible via a specific docker tag.

_Bioconductor_ has several images in addition to 'base', specific to various areas of research.
The 'core' image installs the most commonly used _Bioconductor_ packages.
_Bioconductor_ images for proteomics, metabolomics, and flow cytometry are community maintained.
All community maintained images build on top of the _Bioconductor_ base image and therefore indirectly the Rocker images.
To simplify building and maintaining _Bioconductor_ images, we use a Ruby templating engine.
A recent audit of the _Bioconductor_ Dockerfiles, following best practices from the Docker website, led to a reduction in the size and number of layers.
The most important insights involve the ['union' file system used by Docker](https://docs.docker.com/storage/storagedriver/overlayfs-driver/#how-container-reads-and-writes-work-with-overlay-or-overlay2).
In this file system, once a layer (e.g., `RUN` statement) writes to a file path, the file path is never altered.
A subsequent layer that might appear to remove or overwrite the path actually masks, rather than alters, the original.
It is therefore important to clean up (e.g., cache removal) within each layer, and to avoid re-installing existing dependencies.

@@

As with any other software, each new version of R comes with its share of changes. 
Some are breaking changes, some are not, but the fact is that running a piece of code in a given version might give different s from another version ; which, in some case, can be an issue when it comes to reproducibility.
For example, think about R random seed: starting with `R 3.6.0`, running `set.seed(2811); sample(1:1000, 2)` will not give the same result as if it was run inside an older version of R.
That might seem trivial, but all the code using the random number generator will not be exactly reproducible after this breaking change. 
Detecting this kind of behavior is the reason why the [r-online](https://github.com/ColinFay/ronline) app has been developped: available as a standalone NodeJS app or [online](https://srv.colinfay.me/r-online), this app was designed for historic exploration of R: with it, the user can compare a piece of code run in two separate versions of R. 
Internally, this app takes the piece of code written by the user, opens one or two docker instances with the given version of R, and returns the result to the user.

Containers are perfectly suited to capture a specific configuration of a computing environment to prohibit such problems.
The Rocker images provides the _versioned stack_ for different the version of R since the project inception, which use the _then-current_ stable Debian base image \citep[cf.][]{RJ-2017-065}.
Based on custom build phase hooks\footnote{\href{https://docs.docker.com/docker-hub/builds/advanced/}{https://docs.docker.com/docker-hub/builds/advanced/}}, i.e. small shell scripts executed at different phases in the automated build of Rocker images, there are also semantic version tags for the most recent freezed (i.e. using MRAN) versions with only the major and minor version\footnote{See \href{https://github.com/rocker-org/rocker-versioned/blob/master/VERSIONS.md}{https://github.com/rocker-org/rocker-versioned/blob/master/VERSIONS.md}, FIXME: \href{https://github.com/rocker-org/rocker-versioned/issues/42}{https://github.com/rocker-org/rocker-versioned/issues/42}.}.
So, at the time of writing this article, `rocker/r-ver:3` and `3.6` are aliases for `3.6.0`, because `3.6.1` is the latest release.
With the release of `3.6.2` and pinning of the MRAN version in `3.6.1`, the tags `3` and `3.6` will deliver the same image as `3.6.1`.
These tags allow users to update the base image to retrieve bugfixes while reducing the risk of also introducing breaking changes.

@@

# R Packages
<!-- from packages closer to Docker to second level packages to more complex ones -->

## Clients

A number first order R packages exist that talk directly to Docker.
They can be used to control the Docker daemon, e.g. to build images or run containers, from R without explicit knowledge of the Docker command line interface.

- [`stevedore`](https://github.com/richfitz/stevedore)
- [`docker`](https://bhaskarvk.github.io/docker/)
- [`harbor`](https://github.com/wch/harbor/)
- [`dockermachine`](https://github.com/cboettig/dockermachine)
- [`dockyard`](https://github.com/thebioengineer/dockyard)

## Capture and create environments [@nuest]

Several second order R packages attempt to make the process of creating Docker images and using containers for specific tasks, such as running tests or rendering reproducible reports, easier.

- [`dockerfiler`](https://github.com/ColinFay/dockerfiler/)
- [`containerit`](https://github.com/o2r-project/containerit/)
- [`dockertest`](https://github.com/traitecoevo/dockertest/)

**[_liftr_](https://nanx.me/liftr/)** \citep{liftr2019} aims to solve the problem of persistent reproducible reporting in statistical computing.
Currently, the R Markdown format and its backend compilation engine _knitr_ offer a _de facto_ standard for creating dynamic documents \citep{xie2018}.
However, the reproducibility of such content authoring environments is often limited to individual machines --- it is not easy to replicate the system environment (libraries, R versions, R packages) where the document was compiled.
This issue becomes even more serious when it comes to collaborative document authoring and creating large-scale document building services.
_liftr_ solves this reproducibility problem by bringing Docker to the game.
In essence, _liftr_ helps R Markdown users create and manage Docker containers for rendering the documents, thus make the computations utterly reproducible across machines and systems.

@@

- [`rize` for Shiny](https://github.com/cole-brokamp/rize)

One of the key tool of Docker are dockerfiles, which can be thought of as "recipes" used to build a specific image. 
But building these files can feel like a workflow break as it demands to open a separate file, and / or to write a small wrapper around R's `write()` function to add elements to the file. 
At the same time, doing it by hand prevents from using programming language for what they are good for: iteration, and automation. 
Iteration and automation for Dockerfiles is the very reason behind [`dockerfiler`](https://github.com/ColinFay/dockerfiler), an R package designed for building Dockerfiles straigth from R.

For example, the **[`golem`](https://github.com/ThinkR-open/golem)** package makes an heavy use of `dockerfiler` when it comes to creating the `Dockerfile` for building production-grade Shiny applications and deploying them.
The reason behind this is that `dockerfiler`, on top of being scriptable from R, can leverage all the tools available in R to parse a `DESCRIPTION` file, to get system requirements, to list dependencies, versions, etc.

@@

## Deployment to cloud services [div]

The cloud is the natural environment of containers, and becomes the go-to mechanism to expose applications written in R to users.

- [`babelwhale`](https://cran.r-project.org/web/packages/babelwhale/index.html): Running a Docker from R with Singularity or Docker as back-end. This is really useful in HPC environments where a user might not have root access but is able to install Singularity instead of Docker. [@rcannood]
- `RSelenium`
- [`googleComputeEngineR`](https://cloudyr.github.io/googleComputeEngineR/) (function `docker_run`)
  - I don't think `docker_run` is main focus for containers for this package, its more how it uses Docker to launch custom environments that can be templated (e.g. rstudio, shiny etc.) and for parrallisation e.g. `library(future)`  using GCE VMs as endpoints. [@MarkEdmondson1234]
- [`analogsea`](https://github.com/sckott/analogsea) (digital ocean R client)
- [`AzureContainers`](https://cran.r-project.org/package=AzureContainers): Umbrella package for working with containers in Microsoft's Azure cloud. Provides interfaces to three Azure services: Container Registry, Container Instances (for running individual containers), and Kubernetes Services (for orchestrated deployments)

## Using R to power enterprise software in production environments [@jnolis, @nolistic]

R has been historically viewed as a tool for analysis and scientific research--not for creating software that corporations can rely on to continously run in real time. However, thanks to advancements in R running as a web service, along with with the ability to deploy R in docker containers, modern enterprises are now capable of having real-time machine learning powered by R.

The backbone of this work is the _[plumber](https://www.rplumber.io/docs/hosting.html#docker)_ package, created by Jeff Allen. The plumber package allows R to be used to host web services through RESTful APIs. When R is running on a server, a different server can ask R to make a computation through an HTTP request and get an immediate response. For instance, if a data scientist at a company creates a machine learning model to tell which customers are likely to make a repeat purchase, if they serve the model through a plumber API then someone else on the marketing team could write software that sends different emails depending on that real-time prediction. Plumber syntax is simple to read and understand: the API is created by commenting above function names in an R script.

```
#* Return the sum of two numbers
#* @get /random
function(){
  runif()
}
```
_Example of a plumber API to generated a random number by visiting `http://127.0.0.1/random`_

>>>>

---
title: "Containers and R---the Rockerverse and beyond"
author:
  - name: Daniel Nüst
    affiliation: Institute for Geoinformatics, University of Münster
    address:
    - Heisenbergstr. 2
    - 48149 Münster
    - Germany
    - \orcid{0000-0002-0024-5046}
    email: daniel.nuest@uni-muenster.de
  - name: Carl Boettiger
    #affiliation: -
    #email: -
  - name: Robrecht Cannoodt
    #affiliatin: -
    #email: -
  - name: Dirk Eddelbuettel
    affiliation: University of Illinois at Urbana-Champaign
    address:
      - Department of Statistics
      - Illini Hall, 725 S Wright St
      - Champaign, IL 61820
      - \orcid{0000-0001-6419-907X}
    email: dirkd@eddelbuettel.com
  - name: Mark Edmondson
    affiliation: IIH Nordic A/S, Google Developer Expert for GCP
    email: mark@markedmondson.me
  - name: Colin Fay
    affiliation: ThinkR
    email: contact@colinfay.me
  - name: Ben Marwick
    #affiliation: -
    #email: -
  - name: Karthik Ram
    #affiliation: -
    #email: -
  - name: Noam Ross
    #affiliation: -
    #email: -
  - name: Nan Xiao
    affiliation: Seven Bridges Genomics
    address:
      - 529 Main St, Suite 6610
      - Charlestown, MA 02129, USA
      - \orcid{0000-0002-0250-5673}
    email: me@nanx.me
  - name: Lori Shepherd
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
      - Elm & Carlton Streets
      - Buffalo, New York, 14263 USA
      - \orcid{0000-0002-5910-4010}
    email: lori.shepherd@roswellpark.org
  - name: Nitesh Turaga
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
    - Elm & Carlton Streets
    - Buffalo, New York, 14263 USA
    - \orcid{0000-0002-0224-9817}
    email: nitesh.turaga@roswellpark.org
abstract: >
  The Rocker project provides widely-used Docker images for R across different application scenarios.
  This articles surveys describe downstream projects building upon Rocker.
  We also look beyond Rocker to other projects connecting containerisation with R.
  These use cases and the diversity of applications demonstrate the power of Rocker and containerisation for collaboration, effectivity, scalability, and transparency.
# blank footnote? https://tex.stackexchange.com/questions/170511/footnotes-without-numbering
output:
  rticles::rjournal_article:
    keep_tex: true
---

@@

The R community keeps growing. This can be seen in the number of new packages on CRAN which keeps on rising, but also in the increased number of conferences, online courses, meetups, and unconferences, as well as in the uptake in education and industry increases [REFs].
All this cements the role of R as the _lingua franca_ of statistics, data visualisation, and computational research.
Coinciding with the rise of R was the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distribution and deployment of software.
Combining both these topics, the Rocker Project ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next Section for more details).
The considerable uptake and continued evolution of the Rocker Project has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research to production deployments.
This article presents this _Rockerverse_ of projects across all development stages: demonstrations, prototypes, and mature products.
We also introduce related activities connecting the R language and environment with other containerisation solutions.
The main contributions is a coherent picture of the current lay of the land of using containers in, with, and for R.

# Containerization and Rocker

Docker, an application and service provide by the eponymous company, has in just a few short years risen to prominence for development, testing, deployment and distribution of computer software \citep[cf.][]{datadog_8_2018,munoz_history_2019}.
While there are related approaches such as LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}} or Singularity \citep{kurtzer_singularity_2017}, Docker has become synomymous with "containerization"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operationg systems.
In doing so, Docker had recognised and validated the importance of one very important thread that had been emerging, namely virtualization.
By allowing (one or possibly) multiple applications or services to run concurrently on one host machine without any fear of interference between them, an important scalability opportunity is being provided.
But Docker improved on virtualization by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualization.
This makes for more efficient use of system resources and allowed another order of magnitude in terms of scalability of deployment [REFERENCE!].   
While Docker makes use of Linux kernel features, its importance was large enough so that some required aspects of running Docker have been added to other operating systems to support Docker there more efficiently too \citep{microsoft_linux_2019}.
The success even lead to standardisation and industry collaboration \citep{oci_open_2019}.

The key accomplishment of Docker (as an "application") is to make a "bundled" aggregation of software (the so-called "image") available to any system equipped to run Docker, without requiring much else from the host besides the actual Docker application installation.
This is a rather attractive, and novel, proposition which has lead to widespread adoption and use of Docker in a variety of domains, _e.g._, cloud computing infrastructure \citep{Bernstein2014}, data science \citep{boettiger_introduction_2015}, and edge computing \citep{alam_orchestration_2018} [more examples!].
It provided to be a natural match for "cloud deployment" which runs (or at least appears to run) "seamlessly" without much explicit reference to the underlying machine, architecture or operating system: containers can be deployed with very little in terms of dependencies on the host system---only the container runtime is required.

For statistical computing and analysis centered around R, the Rocker Project has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}.
The Rocker Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with RStudio~Server and Shiny~Server, to domain-specific containers such as `geospatial`.
Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
[Maybe at a later point add forward reference to other topics building on Rocker described here.]

@@

_Microsoft R Open_ ([MRO](https://github.com/nuest/mro-docker)) is an R distribution formerly known as Revolution R Open (RRO) before Revolution Analytics was acquired by Microsoft.
MRO is compatible with main R and it's packages.sss
"It includes additional capabilities for improved performance, reproducibility, and platform support." \citep{microsoft_mro}
Most notably these capabilities are the MRAN repository, which is enabled by default, and the (optional) integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread}.
MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images `nuest/mro` are available on Docker Hub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) and on GitHub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) respectively.
The images are inspired by the Rocker images and can be used much in the same fashion, effectively a drop-in replacement allowing users to quickly evaluate if the benefits of MRO + Intel® MKL apply to their use case.
Version-tagged images are provided for the latest bugfix release of recent R versions.
Extended license information about MKL is printed at every startup.

_Renjin_ is an interpreter for the R language running on top of the [Java Virtual Machine](https://en.wikipedia.org/wiki/Java_virtual_machine) (JVM) providing full two-way access between Java and R code \citep{wikipedia_renjin_2018}.
It was developed to combine the benefits of R, such scripting and extension packages, with the JVM's advantages in the areas of security, cross-platform availability, and established position in enterprise settings.
R extension packages need to be specially compiled and are distributed via the Java package manager [Apache Maven](https://en.wikipedia.org/wiki/Apache_Maven), cf. [http://packages.renjin.org/packages](http://packages.renjin.org/packages) for available packages.
Packages are loaded on demand, i.e. at the first call to `library()`.
Not all R packages, especially one linking to binary libraries, are available, e.g. `rgdal` \footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.

There are no offical Docker images for Renjin, but community-maintained images for selected releases only are available under `nuest/renjin` on Docker Hub and GitHub at [https://hub.docker.com/r/nuest/renjin](https://hub.docker.com/r/nuest/renjin) and [https://github.com/nuest/renjin-docker](https://github.com/nuest/renjin-docker) respectively.
These images expose the command line interface of Renjin in a similar fashion as Rocker images and allow an easy evaluation of Renjin's suitability, but are not intended for production use.

[_pqR_](http://www.pqr-project.org/) is a "a pretty quick version of R".
_pqR_ fixes some opinionated issues in the R language and is the basis for experimental features, e.g. automatic diffentiation\footnote{ \href{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}}.
The source code development [on GitHub](https://github.com/radfordneal/pqR/) is a one man project and it does not provide any Docker images.
But especially disruptive approaches may contribute to the development of the R ecosystem, so the `nuest/pqr` images on Docker Hub and GitHub at [https://hub.docker.com/r/nuest/pqr/](https://hub.docker.com/r/nuest/pqr/) and [https://github.com/nuest/pqr-docker](https://github.com/nuest/pqr-docker) respectively.

_FastR_ is "A high-performance implementation of the R programming language, built on GraalVM" (https://github.com/oracle/fastr).
It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R \footnote{ \href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.

There are no offical Docker images provided, but `nuest/fastr-docker` provides image and `Dockerfile` on Docker Hub and GitHub at []() and [https://github.com/nuest/fastr-docker](https://github.com/nuest/fastr-docker) respectively.

While the images presented in this section are far from being as vetted, stable, and widely used as any of the _Rocker_ images, they demonstrate an important advantage of containerisation technology, namely the abiliy to transparently build portable stacks of open source software and make them easily accessible to users.
All different distributions are published under GPL licenses.
Since all of the different R distributions claim better performance as a core motivation, a comparision based on Docker images, potentially leveraging the [resource restriction mechanisms](https://docs.docker.com/config/containers/resource_constraints/) of Docker to level the playing field, seems useful future work.

@@

[_Bioconductor_](http://bioconductor.org/help/docker/) is an open source, open development project for the analysis and comprehension of genomic data \citep{gentleman_bioconductor_2004}.
The project consists of 1741 R software packages as of August 15th 2019, as well as packages containing annotation or experiment data.
_Bioconductor_has a semiannual release cycle, each release is associated with a particular version of R.
Docker images allow availability of current and past versions of _Bioconductor_ for convenience and reproducibility.

_Bioconductor_ 'base' docker images are built on top of `rocker/r-ver` and `rocker/rstudio`.
_Bioconductor_ installs packages based on the R version, and therefore uses `rocker/rstudio` and `rocker/r-ver` version tagging.
_Bioconductor_ selects the desired version of R from _Rocker_, adds the BiocManager CRAN package for installing appropriate versions of _Bioconductor_ packages, and creates a _Bioconductor_ docker image with an informative tag (R\_version\_Bioc\_version).
The images are summarized on the _Bioconductor_ web site (https://bioconductor.org/help/docker/), maintained on GitHub ([https://github.com/Bioconductor/bioc_docker](https://github.com/Bioconductor/bioc_docker)), and available to the community through [DockerHub](https://hub.docker.com/).
Past and current combinations of R and _Bioconductor_ are therefore accessible via a specific docker tag.

_Bioconductor_ has several images in addition to 'base', specific to various areas of research.
The 'core' image installs the most commonly used _Bioconductor_ packages.
_Bioconductor_ images for proteomics, metabolomics, and flow cytometry are community maintained.
All community maintained images build on top of the _Bioconductor_ base image and therefore indirectly the _Rocker_ images.
To simplify building and maintaining _Bioconductor_ images, we use a Ruby templating engine.

A recent audit of our Dockerfiles, following best practices from the Docker website, led to a reduction in the size and number of layers in images we produce.
The most important insights involve the ['union' file system used by Docker](https://docs.docker.com/storage/storagedriver/overlayfs-driver/#how-container-reads-and-writes-work-with-overlay-or-overlay2).
In this file system, once a layer (e.g., `RUN` statement) writes to a file path, the file path is never altered.
A subsequent layer that might appear to remove or overwrite the path actually masks, rather than alters, the original.
It is therefore important to clean up (e.g., cache removal) within each layer, and to avoid re-installing existing dependencies.

@@

- run code with Versions of R: https://srv.colinfay.me/r-online
- old R versions: https://github.com/rocker-org/rocker-versioned/issues/138
- [semantic version tags for Rocker](https://github.com/rocker-org/rocker-versioned/#version-tags)
- debugging memory problems across multiple versions: https://github.com/wch/r-debug
- cross-R testing: Show how you can run the same script in multiple R variants in containers (new contribution). Ideally the images should be created using `containerit` and controlled with `stevedore`/`docker`, i.e. via an R script.

@@

# R Packages

## Packages for automation and packaging [@nuest]

- [`containerit`](https://github.com/o2r-project/containerit/)
- [`dockertest`](https://github.com/traitecoevo/dockertest/)

### liftr

[_liftr_](https://nanx.me/liftr/) \citep{liftr2019} aims to solve the problem of persistent reproducible reporting in statistical computing.
Currently, the R Markdown format and its backend compilation engine _knitr_ offer a _de facto_ standard for creating dynamic documents \citep{xie2018}.
However, the reproducibility of such content authoring environments is often limited to individual machines --- it is not easy to replicate the system environment (libraries, R versions, R packages) where the document was compiled.
This issue becomes even more serious when it comes to collaborative document authoring and creating large-scale document building services.
_liftr_ solves this reproducibility problem by bringing Docker to the game.
In essence, _liftr_ helps R Markdown users create and manage Docker containers for rendering the documents, thus make the computations utterly reproducible across machines and systems.

@@

- [`rize` for Shiny](https://github.com/cole-brokamp/rize)
- [`dockerfiler`](https://github.com/ColinFay/dockerfiler)

## Packages for control and provisioning [@nuest, @richfitz, @rcannood]

- [`stevedore`](https://github.com/richfitz/stevedore)
- [`babelwhale`](https://cran.r-project.org/web/packages/babelwhale/index.html): Running a Docker from R with Singularity or Docker as back-end. This is really useful in HPC environments where a user might not have root access but is able to install Singularity instead of Docker. [@rcannood]
- [`docker`](https://bhaskarvk.github.io/docker/)
- `RSelenium`
- [`googleComputeEngineR`](https://cloudyr.github.io/googleComputeEngineR/) (function `docker_run`)
  - I don't think `docker_run` is main focus for containers for this package, its more how it uses Docker to launch custom environments that can be templated (e.g. rstudio, shiny etc.) and for parrallisation e.g. `library(future)`  using GCE VMs as endpoints. [@MarkEdmondson1234]
- [`analogsea`](https://github.com/sckott/analogsea)
- [`harbor`](https://github.com/wch/harbor/)
- [`dockermachine`](https://github.com/cboettig/dockermachine)

@@
